

\section{General Discussion}

The central aim of this thesis was to examine the measurement of treatment
expectancies in the context of the placebo effect. This thesis used self-report,interview, implicit and physiological variables to deliver greater understanding of these psychological constructs and their relationship with the placebo effect.


The thesis was structured into a number of parts, each of which will be discussed in turn, setting the Discussions at the end of each chapter into a fuller context. 

The first section of the thesis was the review of the literature around the placebo effect and implicit measures, along with any known individual-level predictors of the effect. This review found that placebo effects have been seen to be mediated by optimism (as operationalised by the Life Orientation Test, Revised (LOTR) \cite{Scheier1994}) \cite{Geers2005,morton2009reproducibility} in both pain and non-pain paradigms. This review also linked the constructs of optimism, which can be defined as ``generalised outcome expectancies around the future'' \cite{Carver2010}, with the construct of response expectancies of Kirsch \cite{Kirsch1985,Kirsch1997}, which are described as the ``expectation of a non-volitional response''. These constructs both crucially rely on the participants' perception of their likelihood, and a secondary aim of the thesis developed which was to examine the usefulness of these as two separate rather than one combined predictor. The hypothesis was that these two constructs should be significantly correlated, and that one of them would mediate the impact of the other on the observed placebo response \cite{Geers2005}. 

Another important potential moderator variable came to light in the literature review, which was the construct of mindfulness, as operationalised by the Mindful Attention Absorption Scale (MAAS) \cite{brown2003benefits}. This variable was found to moderate the correlations between explicit and implicit measures in an experience sampling study. Furthermore, implicit measures were found to be better predictors of spontaneous responses, while explicit measures were found to be better predictors of deliberate behaviour \cite{Levesque2007}, a pattern known as the double-dissociation effect and observed in IAT studies in a variety of domains \cite{Asendorpf2002,Perugini2005,Grumm2007,Steffens2006}. 

This research literature suggested that a good strategy would be to replicate the LOT-R and MAAS on a number of samples from which the eventual experimental sample would be drawn, both to develop better models for these constructs' relationship (as they had not been concurrently examined when this work was carried out) and to tailor the measures towards the experimental sample, a strategy which was hoped to increase the precision of estimation on the overall sample. In this research, the RAND-MOS was used both to replicate the impact of these primary constructs on health, and to provide a weak link to health and treatment expectancies in preparation for the experimental research. 

One large problem with the development of an implicit measure of treatment expectancies was that there existed no gold-standard measure of explicit treatment expectancies, despite them having been shown to materially affect the outcome of clinical trials \cite{Linde2007,Bausell2005,Benedetti2005}. This was an issue both because typical methods of IAT development rely on such a self-report measure, and because without such a measure, the incremental improvement (if any) granted by an explicit measure would not be clear. Therefore, the development and testing of such a measure formed part of the work of this thesis. 

Another fact which became extremely clear during the course of the literature review, is that current methods of both developing and validating implicit measures (or more specifically, IAT's), rely extremely heavily on the existence of prior self-report instruments. For the placebo effect which was the focus of primary investigation, no such self-report measure existed, which meant that a new approach needed to be tested. A review of methods in this space revealed the work of George Kelly on personal construct theory \cite{Kelly1991}, and more specifically the repertory gird introduced in Book 1 of the aforementioned reference. It was believed that such an approach might prove useful in the current case, and might also prove useful with other constructs of interest in the future. \footnote{I am indebted to Dr's Jurek Kirakowski and Sean Hammond for originally suggesting this idea.}

Therefore, in order to develop these stimuli, it was decided to do some qualitative research around health constructs with doctors, alternative therapists and students, to gain an appreciation for how they contextualised and understood issues around health, medicine and sickness. The results of this procedure, along with a parallel exercise asking participants to rate the five most important people or qualities of people related to health, would then be used to develop the Health Repertory Grid. 

Finally, the development of the experiment was planned, and it was decided to use a placebo analgesia design, inducing pain with the sub-maximal torniquet technique \cite{moore1979submaximal}, as this form of pain induction has been shown in a meta-analysis to produce the largest effect sizes \cite{Sauro2005}. 

Below, the results and findings from each of these sections are described and placed in a context of larger research. 

\section{Health, Optimism and Mindfulness}
\label{sec:health-optim-mindf}

This study consisted of approximately fifteen hundred (N=1500) observations, of which 1100 were collected in an online fashion, and four hundred were collected in pen and paper format. 

The major aims of this part of the research was to collect useful background data for the experimental sample, enabling psychometric models to be built on large samples, which could then be applied to the experimtal sample, as well as developing psychometric models which could tease out the relationships between the health-related constructs (RAND MOS), optimism (LOT-R) and mindfulness (MAAS). The headline finding of this research, which was replicated across both samples (and a further one which was not analysed) was that optimism was negatively related to both health and mindfulness, a finding which has not been reported before. 

This finding can be better understood in the light of some of the SEM results on the same dataset. In these, it was determined that emotional well-being moderated the impact of this suprising effect. 


Additionally, this strange phenomenon replicated across both original studies, and also in the pilot experiment. However, it did not replicate in the main experimental study, raising the issue of what actually occurred here differently. Upon examination of the materials, the only difference between the setup for the experimental study was the order in which the measures were administered. In all the previous studies, the MAAS was filled out before the LOT-R, while in the experimental study, the order was opposite. 

This finding links in with the other finding that emotional well-being (a subscale of the RAND-MOS) moderated the effect of optimism on health. To see this, merely look at the questions which make up the Emotional Well-Being scale, and note its similarity to the LOT-R. It appears that what occurred is that participants who would typically respond highly to optimism (as evidenced by their Emotional Well-Being score) had their response patterns altered by the act of completing the MAAS. This occurred because the MAAS operationalises mindfulness in terms of negative mindlessness. This may have caused the participants high in Optimism to have realised that some of their optimism was unjustified, which therefore depressed their responses on the LOT-R, leading to the unexpected negative correlation with health. However, this is somewhat circumstantial given that this hypothesis was not directly tested in this research. 

Additionally, the cross-validation approaches employed in this section of the research showed an interesting pattern. The factor models which were developed using the reduced scales from IRT analyses showed better predictive ability on unseen data than did those which were developed from factor analysis. This suggests that an interesting approach to the development of scales would be to use IRT (specifically Mokken analyses) to reduce the number of items, and use this reduced scale in SEM to determine if this effect is widespread or merely circumstantial to the samples collected here. 

\section{Explicit Expectancy Measure}
\label{sec:expl-expect-meas}

The next step in the research was the development of an explicit expectancy measure, which would capture the multi-dimensional nature of treatment expectancies more fully \cite{Stone2005}. Again, this measure was developed across two samples, a small first sample which aimed to validate the measure in terms of reliability, and a second larger study which incorporated the results of the first study and utilised the Beliefs About Medicine Questionnaire (BAMQ) \cite{Horne1999} to provide evidence of convergent validity for the measure. 

The results of the first study showed that the measure had extremely good split-half reliability ($ \alpha=0.9$). Another finding was that the first instrument broke down in factor analytic terms in terms of Conventional (Pills, Creams, Injections) and Alternative (Acupuncture) treatments. This was obvious both from the intercorrelations of each of the questions, and from the results of factor analytic and IRT modelling. This suggested that two other alternative treatments should be added to the questionnaire, to determine if this was a valid structure and to provide balance. 

The second study showed the expected correlations  with the BAMQ measure (positive for alternative treatments, negative for conventional treatments) which suggested that the measure did have convergent validity. Additionally, the notion of conventional and alternative treatments being separate factors was replicated across both samples, and using factor analytic and IRT methods. Again, the backtesting strategy employed in the previouc chapter, using a combination of IRT and FA approaches, performed better than either of these approaches alone, suggesting that this may be a useful methodology in general. 

%%insert some interesting SEM results here, when available. 

\section{Qualitative Research, Rep Grids and Experimental Pilot}
\label{sec:qual-rese-rep}

The final chapter before the main experimental approach was the analysis of the qualitative data, the development of the repertory grid and the development of the IAT measures and their piloting. 

From the coding and analysis of the interviews, it was determined that the major contrast between styles of healthcare (conventional and alternative) was that the alternative therapists regarded themselves as treating a person, whereas the GP's focused more on the symptoms which this person presented with. Additionally, the GP's seemed to focus more on external causes (pathogens, environment) whereas the alternative therapist focused more on internal causes (body-mind linkages, thought patterns). Note that both of the groups focused on the nature of health as both a social and personal responsibility, but that they disagreed on where the responsibility for changes lay. The students in the sample provided either a moderately alternative viewpoint, or a moderately conventional viewpoint, but in no case were they as extreme as any of the practitioners. The major differential theme appeared to be that of conventional versus alternative, and the truth or falsity of health claims, and this was incorporated into the development of the Health Repertory Grid (and later into the design of the Treatment Credibility IAT). 

The next step was the development of the Health Repertory Grid. This was developed analogously to the original repertory grid, using health related constructs taken both from the interviews and from the questionnaire asking people to name the five most important health-related people or qualities in their life. Unfortunately, it became apparent in the course of piloting this instrument that participants did not have enough examples of these constructs to enable them to complete this task. Therefore, the treatment credibility IAT was developed using the notion of true versus false and conventional treatments versus alternative treatments, while the optimism IAT was developed from the LOT-R, in line with current practice in the field. 

Finally, the two implicit measures were administered to a small sample of voluinteers, and showed the expected correlations (small, but in the right direction) with the explicit measures of the same constructs. Additionally, when the experiment was piloted, there was a small effect of the IAT's on the probability of placebo response in a logistic regression. 

\section{Experimental Research}
\label{sec:exper-rese}

The final part of the thesis was the experimental research. One hundred and eleven (N=111) participants were assigned to either a Deceptive (told drug, got placebo), Open (told placebo, got placebo) or No Treatment Condition. Additionally, participants were randomised to either a Prime or No Prime condition. The major finding in terms of condition was that those participants who were in the Open condition showed a greater placebo response than those who were in the Deceptive Condition. This finding was extremely unexpected, and has not been seen in the literature before. However, given that the suggestions focused on ``clinically proven'' (in the Open condition) and ``recently approved'' in the Deceptive condition, this may have had an impact. Additionally, there was a significant Prime by Condition interaction, suggesting that priming was differentially impacted by by the condition (Open Placebo participants showed a greater propensity to respond to suggestion after being primed). This would seem to have potentially large implications for medical practice, if replicated. 

With respect to the major hypothesis of the thesis, no direct impact of the explicit or implicit measures was shown to be significant in a logistic regression. However, a three way interaction was shown to be significant, however, this model suffered from issues of multi-collinearity and so should not be taken as accurate without replication. Attempts were made to utilise machine learning techniques for the prediction of placebo response, and while some of these techniques were extremely effective at predicting the lack of a placebo response, none performed at above chance level at predicting positive placebo response. However, it would be useful in terms of the design of clinical trials if participants who would reliably respond to placebo could be eliminated, and so this measure might prove useful if replicated. 

Finally, the relationship between skin conductance and pain ratings was examined. This analysis showed that for the two treatment conditions, entirely different GSR responses were shown. This would seem to indicate that the difference between certain and uncertain expectancies can be examined using simple GSR equipment as opposed to expensive fMRI equipment. 

In conclusion, there was no significant relationship shown between implicit measures and placebo in standard analyses, but there were some indications that a machine learning approach might be able to use the measures developed in this thesis to retain participants who were unlikely to respond to placebo. 

\section{Conclusions and Further Research}

%%% Local Variables:
%%% TeX-master: "PlaceboMeasurementByMultipleMethods"
%%% End: